---
title: "Using plasticRank"
format: html
vignette: >
  %\VignetteIndexEntry{Using plasticRank}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  collapse: true
  comment: "#>"
---

```{r setup}
library(plasticRank)
```

## Overview

**plasticRank** provides a streamlined pipeline for loading, cleaning,
summarizing, and visualizing a merged dataset of plastic waste audit data
(TidyTuesday 2021) joined with GDP, population, and CO2 emissions data.
The dataset covers 2019–2020 across roughly 80 countries.

The typical workflow is straightforward:

1. Load the raw data
2. Clean it
3. Pass the clean dataset into whichever functions you need

---

## 1. Load the data

`load_data()` reads the dataset bundled with the package. No internet
connection required.

```{r}
df <- load_data()
dim(df)
```

---

## 2. Clean the data

`clean_data()` removes the ~48,000 "EMPTY" audit records that have no country
attached, drops Taiwan entries from the country code lookup, and removes rows
where every per-capita column is missing. The result is what you pass to
everything else.

```{r}
df_clean <- clean_data(df)
dim(df_clean)
```

---

## 3. Summarize by country

`country_summary()` rolls up all audit entries to one row per country for a
chosen year.

```{r}
country_summary(2019, data = df_clean)
```

---

## 4. Rank countries

`rank_table()` ranks every country from highest to lowest for a chosen variable
and year.

```{r}
rank_table("co2_per_capita", 2019, data = df_clean)
```

---

## 5. Bubble plot

`bubble()` plots two variables against each other, with bubble size showing
a third. The log transform helps when the data is heavily skewed.

```{r}
bubble(
  x           = "gdp_per_capita",
  y           = "co2_per_capita",
  size        = "population",
  transform   = "log",
  year_select = 2019,
  data        = df_clean
)
```

---

## 6. Bar chart of top countries

`all_ranks()` shows the top countries for a variable as a horizontal bar chart.

```{r}
all_ranks("plastic_waste_per_capita", year_select = 2019, top_n = 15, data = df_clean)
```

---

## 7. Statistical analysis

`calc_correlation()` runs a Pearson correlation test between two variables and
returns the coefficient, p-value, and confidence interval.

```{r}
calc_correlation("gdp_per_capita", "co2_per_capita", data = df_clean)
```
